Interactive Visual Analysis of Image-Centric Cohort Study Data

### 2014-06-18 draft.tex
Added new figure for application part and finished it

The front-end is created using \texttt{Bootstrap}\footnote{Developed by Twitter, \url{http://getbootstrap.com}} as foundation for the layout and basic UI elements using \texttt{HTML5}, \texttt{CSS3} and \texttt{Javascript}.
%
Information visualizations such as scatterplots and bar charts are created using the popular \texttt{Data-Driven Documents (D3.js)} library \cite{D3}, which works well for attaching data to visible elements like vector graphics and provides powerful transformation and mapping tools.
%
The pivot table implementation was adapted using \texttt{PivotTable.js}\footnote{Developed by Nicolas Kruchten, \url{http://nicolas.kruchten.com/pivottable}}.
%
\texttt{Three.js}\footnote{Originally developed by Ricardo Cabello, \url{http://threejs.org}} allows GPU-accelerated data rendering using \texttt{WebGL}.
%
Communication between client and server is enabled via the WebSockets protocol.
%
Since our clustering algorithms are written in \texttt{MatLab}\footnote{Owned by The MathWorks, \url{http://mathworks.com}}, we had to access them using the \texttt{NodeJS} server.
%
We accomplished this by converting it to a parameterized standalone console application that is spawned by \texttt{NodeJS} on client request and then reads the result from the console standard out and returns it in a proper format to the client.
%
All parameter-steered console applications can be incorporated in this context.
\\\\
\section{Application}
This section describes how our presented \emph{IVA} workflow is used for an epidemiological use case.
%
We conducted a case study with two domain experts on a data set compiled to analyze lower back pain. 
%
This is one of the most common reasons for an adult to see a physician in the Western civilization \cite{Backpain}.
%
Epidemiological analysis of lumbar back pain, such as the work of Harreby et al. \cite{Harreby1996}, is largely focused on non-image information.
%
In comparable studies, only a few shape-related features are included \cite{Lang2011}.
%
To our knowledge, this is the first approach, where shape-related information of the whole lumbar spine is analyzed along with other epidemiological features.
%
Determining risk factors in this area can lead to \cite{Fletcher2012}:
\begin{itemize}
	\item a better understanding of effects of preventive measures such as occupational health and safety regulations,
	\item prognostic features for diagnosis and treatment of lumbar back pain, and
	\item determination of particularly affected risk groups.
\end{itemize}
%
Characterizing the healthy aging process of the spine is a long-term goal for determining age-normalized probabilities for spine-related diseases by incorporating individual risk factors.
%
\subsection{The Lumbar Spine Data Set}
We divide the data set into image and non-image data.
%
There are 127 features describing diagnosed diseases, lifestyle factors, women-specific factors, pain indicators, laboratory values and somatometric features.
%
The image data was acquired on a 1.5~Tesla scanner (Magnetom Avanto; Siemens Medical Solutions, Erlangen, Germany) by four trained technicians in a standardized way.
%
%The spine protocol consisted of a sagittal T1-weighted turbo-spin-echo sequence (676 / 12 [repetition time msec / echo time msec]; 150$^\circ$ flip angle; 500~mm field of view; $1.1\times1.1\times4.0~mm$ voxels) and a sagittal T2-weighted turbo-spin-echo sequence (3760 / 106 [repetition time msec / echo time msec]; 180$^\circ$ flip angle; 500~mm field of view; $1.1\times1.1\times4.0~mm$ voxels) \cite{Hegenscheid2013}.
The spine protocol consisted of sagittal T1- and T2-weighted turbo-spin-echo sequences with rather large slice distance only sufficient for characterizing the lumbar spine \cite{Hegenscheid2013}.

@@

\subsubsection{Shape Visualization and Clustering}
%
The tetrahedron-based detection model described in Section~\ref{application:Data Preprocessing} consists of corresponding grid points for each structure instance.
%
This allows for calculation of shape distance and similarity.
%
This information is used to calculate mean shapes as described in Section~\ref{Interaction- and Visualization Techniques}.

Shape distance is mapped onto color.
%
For dichotomous variables, the color codes distances between mean shapes of the two groups, for variables with more than two manifestations it encodes the distance to the global mean shape of all subjects (see Figure~\ref{fig:hypothesisfree}).
\\\\
Shape-based clustering is carried out via agglomerative hierarchical clustering of the spine canal centerlines (recall Section~\ref{application:Data Preprocessing} and \cite{Klemm2013VMV}).
%
Since the ``correct'' number of clusters in a given group is unknown, an estimate is computed by means of the knee/elbow method \cite{Salvador2004}.
%
The method has proven to produce sound results on a preliminary data set and was able to reproduce textbook knowledge \cite{Klemm2013VMV}.
\subsection{Domain Experts, Setup and Procedure}
%
%Lam et al. divide evaluations of Visual Analysis systems in two categories, 
%
%\com{Anmerkung von Steffen einbauen: Evaluierung geht um fachliches Feedback - Wissen über epidemiologie UND rücken und wissen um die Daten - deswegen schwer ganz viele Experten ranzuziehen}
Guided by the work of Lam et al \cite{Lam2012}, we conducted an investigation of \emph{Visual Data Analysis and Reasoning} (\emph{VDAR}).
%
This approach aims to characterizes the ability of a system to explore data, discover knowledge, generate hypotheses and help formulating decisions.
%
Since it is hard to quantify when these goals are achieved, Lam et al. suggest case studies for the \emph{VDAR} analysis by applying the think aloud protocol to understand the domain experts observations, inferences and conclusions when using the system.

We conducted a qualitative case study with two epidemiological domain experts who also co-authored this publication.
%
Both of them are practicing physicians with focus on epidemiological research.
%
One is an specialist in internal medicine and also the head of the SHIP, the other an radiologist and responsible for the MRI-data acquisition of the entire SHIP. %and has a deep understanding on the available image data.
%
(Their expertise and experience with the data in their day to day research make them aware of the problems associated with the analysis of medical cohort study data.)
%
\paragraph{Setup.} To bypass the large geographical distance between us and the domain experts and to take their busy schedule into account, the case study was done completely web based.
%
Since one major reason for the focus on web technologies, no software needed to be exchanged.
%
A website link was send to the users, which they could open with their browser of choice and start using the tool.
%
To make the user input comprehensible, we observed their monitor via screen sharing.
%
Communication was enabled via webcam supported voice over ip.
%
The total setup time including installing the screen sharing application was about five minutes.
%
Video-recordings of the sessions made it possible to evaluate the sessions later on.
%
\paragraph{Procedure.}
At first, we controlled mouse and keyboard of the participants pc and demonstrated the basic functionalities of the system.
%
This included the Heatmap, the correlation view, how to introduce parameter visualization, how the color coding of differences works and the pivot tables.
%
As they understood the concepts, we handed over the mouse and keyboard control and only observed from this point on.
%
The epidemiologists were given two tasks: one hypothesis-free analysis of the data and one starting with an assumption.

% \subsection{Exploratory Analysis of the Lumbar Spine Data Set}
\subsection{Case 1: Hypothesis-free analysis}
Analyzing the data set without prior hypothesis requires a starting point giving an overview over the data first \cite{Shneiderman1996}.
%
With our tool, there are two ways to achieve this.
%
Performing a \emph{local investigation} by clustering all subjects or use the contingency matrix for a \emph{multivariate analysis}.

\paragraph{Analysis 1.}
%
The latter was chosen first by both domain experts.
%
%They pointed out the high potential they see in this visualization, because it allows to look at the data in an unbiased way.
%They pointed out the benefits of the possibility to look at all variables in the context of each other.
Before, they were not able to to look at all variables in the context of each other.
%
%
To cite one expert, it "illumunates the data black box", making it possible to look at the data unbiased from assumptions.
%
The radiologist was looking for correlations with shape related features in the data, finding that they correlate with \emph{leg pain}, \emph{age}, \emph{body height} and \emph{hormone replacement therapy status}.
%
The internist inspected non-image-derived features, such as age-associated parameters like \emph{income}, \emph{blood fat values} or \emph{number of born children}.
%
This also shows, that even though the visualization may treat every feature equally, the domain experts is more likely interested in features they are familiar with and have personal clinical experience with.
%
Due to the dense mapping of information in the contingency matrix it was suggested to make this visualization full screen.
%Many correlations observed in this view

After this initial overview, the expert performed a \emph{multivariate analysis} by introducing features, such as \emph{age}, \emph{waist circumference}, \emph{weight} or \emph{lumbar spine canal curvature} as bar charts views into the canvas area and selected subgroups to see how they are distributed and if they could observe unusual behavior in the mean-shapes.
%
This pointed out problems with the used categorization method splitting numerical variables into equally sized bins.
%
If an feature contains outliers, such as waist circumference (e.g. subjects with morbid obesity) this approach leads to sparse categories, making it hard to calculate associations.
%
Solutions for this were discussed with the domain experts and are described in more detail in Section~\ref{Lessons Learned}.
%
%\com{Allgemein: Workflow noch einmal grob beschreiben - Overview first, details on demand}
%
\paragraph{Analysis 2.}
\begin{figure}[htb]
 \centering
 %\includegraphics[width=1.5in]{figures/InteractionLoop}
 \includegraphics[width=3.5in]{figures/hypothesisfree}
 \caption{(a) Clustering result of all subjects. The bar chart height indicates the number of subjects in the cluster. The difference to the mean shape is color-coded, whereas blue represents no difference and red a large difference. (b) A pivot table relates cluster results to gender and employment status. Cluster 8 is striking because of the high proportion of women, while cluster 2, 3, 4 contain almost only men.}
 \label{fig:hypothesisfree}
\end{figure}
\com{ToDo: Rewrite}
%
The internist started also with a \emph{multivariate analysis} using the contingency table, but found no associations he was interested in.
%
Therefore, he applied the \emph{local investigation} pattern by a shape grouping step using shape-based clustering via dragging the "All subjects" feature into the canvas area, triggering the shape clustering.
%Shape-based hypothesis-free exploration starts with the \emph{local investigation} pattern by a shape grouping step using shape-based clustering.
%
The results for this step can be seen in Figure~\ref{fig:hypothesisfree} (a).

Cluster 4 represents subjects with average shape.
%
Other shapes differ with respect to size, such as cluster 2, 3, 7, where the last one and cluster 5 also represent a more straight spine, which is usual for subjects with larger body size.
%
Cluster 1 and 6 contain outliers, characterized by their unusual shape and small number.
%
%Cluster 8 has the second largest number of elements and was therefore of special interest.
%
%To perform an automatic feature suggestion using a \emph{multivariate analysis} by looking at \emph{Cram\'{e}r's V} contingency values of in the contingency pane.
To get an overview over the suggested features, the user opened the contingency pane to perform a \emph{multivariate analysis} by looking at \emph{Cram\'{e}r's V} contingency values of all clusters, revealing a strong correlation with \emph{gender} and \emph{body size}.
%
Therefore another \emph{multivariate analysis} was carried out by the feature \emph{gender} to the canvas from the contingency pane and selecting all female subjects (Fig.~\ref{ToDo}).
%
Cluster 1 contained almost only female subjects.
%
Contingency values for this cluster revealed correlations with \emph{leg fatigue}, \emph{physical heavy work}, \emph{body weight}, \emph{dyspnoea} and \emph{headache intensity}.
%
%This revealed associations of this group with employment status, gender, body size, age, thyroid nodules and blood fat value.
%
Since it is an pain indicator, headache was of special interest and was further investigated by incorporating a pivot table setting \emph{headache intensity} in relation to cluster affiliation.
%
Cluster 1 subjects report a much higher chance for heavy headaches than other subjects.
%Another \emph{multivariate analysis} using a pivot table set gender and employment status in relation to clustering affiliation\add{.} 
%
%\add{It s}hows that cluster~8 contains mostly women and also has a larger unemployment rate (see Figure~\ref{fig:hypothesisfree}), while the overall employment rate of women and men in the data set is almost exactly 50\%.
%
%While all observed features seem to be plausible associations related to back pain, the values indicate that cluster~8 contains subjects with chronic back pain radiating to the legs.
%
%Metabolic parameters, such as blood fat and blood sugar, are also possibly associated features.
%
%The employment status is a feature relating to many different lifestyle factors such as income or nutrition as well as age and might act as confounder.
\\\\
The experts emphasized the importance of methods providing overview over the data for hypothesis generation.
%
With the presented \emph{IVA} approaches they were quickly able to confirm medical knowledge as well as elaborate new hypothesis.
% TODO: Contingency Matrix or Adjecency Matrix
%\com{00.44 Clustering-Ergebnis Weiblich und Arbeitspensum Alter}

\subsection{Case 2: Hypothesis-based analysis}
If the user proposes a hypothesis about a relation between a non-image feature regarding shape, the workflow slightly differs from the hypothesis-free analysis.
%
The starting point follows the \emph{feature investigation} pattern, where a feature of interest is selected by dragging it into the canvas area and viewing the subject's distribution as well as their shape differences.
% \paragraph{Analysis 1.}
% The radiologist performed a \emph{feature investigation} by dragging the age variable into the canvas area, observing that age-bins 36-52 years and 52-68 years are nearly identical to the global mean, while 20-36 years and 68-84 years differ in size.
% %
% Performing a \emph{multivariate analysis} by selecting only female subjects via the additionally added gender feature it was observed, that distance was closest to subjects of 68-84 years old and largest to 20-36 years old.
% %
% The expert argued that this is most likely due to the shrinking of the spine with increasing age, making it more similar to the female average spine which is smaller than the global average \com{(00.29)}.
% %
% 
\paragraph{Analysis 1.}
Hypothesis: ``\emph{Back pain is associated with age and lumbar spine shape}''.
%Hypothesis: "Back pain is associated with age and therefore there differences in the lumbar spine shape are observable." 
\com{TODO Image Mosaic Plot Age, Backpain}.
%
To validate this hypothesis, a \emph{multivariate analysis} was performed by introducing the dichotomous feature ``\emph{Did you experience back pain in the last three month?}'' together with the age as mosaic plot by dropping both features on the canvas area.
%
The experts were not able to observe the expected effect in the visualization.
%
Reasons for this are twofold.
%
For one, age influences the lumbar spine shape very heavily, as seen in Fig~\ref{ToDoA}, while the differences between subjects with and without back pain are small.
%
The major difference seen in the visualization are therefore presented by the age feature, masking differences via the back pain parameter.

The second explanation is the commonality of back pain in our society.
%
As seen in Fig~\ref{ToDoA}, subjects reporting back pain are even the majority, which makes it difficult to extract parameters which reliably describe back pain.
%
A \emph{local investigation} using the contingency table showed a strong association between back pain with gender and body height.
%
Body height was explained to be a confounder for gender, since female subjects smaller on average than male subjects.
%
The analysis solely based on shape only accentuated gender differences, not in back pain.
%This makes it hard to analyze shape related features for this disease
%
The next step is to use a variable describing the strength of back pain and only analyzing strong back pain!

The epidemiologists pointed out that they would like to see a more intuitive and fast way to select subgroups based on different features to make full use of the analysis capabilities.
%
We discuss this in detail in Section~\ref{Lessons Learned}.
%gender, use of pain medications. this would be good for future work, providing a better way to subdivide the data set and construct groups. The presented method is very good to gain an overview

@@

% ---------------------------------------------------------------------------------------------------------
\subsection{Further Feedback and Lessons Learned} \label{Lessons Learned}
%\com{ToDo: Lessons Learned der einzelnen Cases - Fokus bei Hypothesenfreier Analyse muss etwa eine gute Übersichtsvisualisierung sein, da es genau das ist was den Epidemiologen hierzu im Moment fehlt}
%
Both domain experts concluded positively over the benefits of the \emph{IVA} approach when analyzing image based cohort study data.
%
One expert emphasized the way the image data is included into information visualizations which comes much more natural to her due to her background in radiology.
%
%Both experts quickly understood the basic concepts of the visualizations and were able to 
%
The domain experts see another great potential in communicating insights efficiently using the presented visualizations \com{(1.08)}.
%
\paragraph{Multivariate analysis is most important for hypothesis generation.}
Both experts emphasized the potential of the \emph{multivariate analysis} capabilities of the adjacency matrix for getting insight into a large amount of features simultaneously.
%
It is also useful to verify established but still controversial risk factors, such as the metabolic syndrome for coronary heart disease and whether the data set provides even more suitable risk factors \com{(39.32)}.
%
Creating adjacency matrices for sub groups, such as different age bins can help to characterize the aging process by deriving age-specific risk factors.

@@

The experts also emphasized that associations related to shape strongly depend on the segmentation quality.
%
The lumbar spine model used in this case study captures deformation of the spine canal well, but lacks precise definition in vertebrae height and shape.
%
Since deformation of the spine canal is commonly the last pathological stage of pathological lumbar spine deformation and is preceded by vertebrae deformation, the system would strongly benefit from more precise segmentation results capturing these prior changes.
%
For the visual comparison itself, the radiologist proposed an abstraction the representation them into representative land marks, such as centers of the vertebrae and cardinal points of the lumbar spine canal.

>>>>

The front-end is created using \texttt{Bootstrap}\footnote{Developed by Twitter, \url{http://getbootstrap.com}} as foundation for the layout and basic UI elements using \texttt{HTML5}, \texttt{CSS3} and \texttt{Javascript}.
%
Information visualizations such as scatterplots and bar charts are created using the popular \texttt{Data-Driven Documents (D3.js)} library \cite{D3}, which works well for attaching data to visible elements like vector graphics and provides powerful transformation and mapping tools.
%
The pivot table implementation was adapted using \texttt{PivotTable.js}\footnote{Developed by Nicolas Kruchten, \url{http://nicolas.kruchten.com/pivottable}}.
%
\texttt{Three.js}\footnote{Originally developed by Ricardo Cabello, \url{http://threejs.org}} allows GPU-accelerated data rendering using \texttt{WebGL}.
%
Communication between client and server is enabled via the WebSockets protocol.
%
Since our clustering algorithms are written in \texttt{MatLab}\footnote{Owned by The MathWorks, \url{http://mathworks.com}}, we had to access them using the \texttt{NodeJS} server.
%
We accomplished this by converting it to a parameterized standalone console application that is spawned by \texttt{NodeJS} on client request and then reads the result from the console standard out and returns it in a proper format to the client.
%
All parameter-steered console applications can be incorporated in this context.
\\\\
\section{Application}
This section describes how the presented \emph{IVA} workflow is used in epidemiological application.
%
We conducted a qualitative case study with two domain experts on a data set compiled to analyze lower back pain. 
%
This is one of the most common reasons for an adult to see a physician in the Western civilization \cite{Backpain}.
%
Epidemiological analysis of lumbar back pain, such as the work of Harreby et al. \cite{Harreby1996}, is largely focused on non-image information.
%
In comparable studies, only a few shape-related features are included \cite{Lang2011}.
%
To our knowledge, this is the first approach, where shape-related information of the whole lumbar spine is analyzed along with other epidemiological features.
%
Determining risk factors in this area can lead to \cite{Fletcher2012}:
\begin{itemize}
	\item a better understanding of effects of preventive measures such as occupational health and safety regulations,
	\item prognostic features for diagnosis and treatment of lumbar back pain, and
	\item determination of particularly affected risk groups.
\end{itemize}
%
Characterizing the healthy aging process of the spine is a long-term goal for determining age-normalized probabilities for spine-related diseases by incorporating individual risk factors.
%
\subsection{The Lumbar Spine Data Set}
We divide the data set into image and non-image data.
%
There are 127 features describing diagnosed diseases, lifestyle factors, women-specific factors, pain indicators, laboratory values and somatometric features.
%
The image data was acquired on a 1.5~Tesla scanner (Magnetom Avanto; Siemens Medical Solutions, Erlangen, Germany) by four trained technicians in a standardized way.
%
%The spine protocol consisted of a sagittal T1-weighted turbo-spin-echo sequence (676 / 12 [repetition time msec / echo time msec]; 150$^\circ$ flip angle; 500~mm field of view; $1.1\times1.1\times4.0~mm$ voxels) and a sagittal T2-weighted turbo-spin-echo sequence (3760 / 106 [repetition time msec / echo time msec]; 180$^\circ$ flip angle; 500~mm field of view; $1.1\times1.1\times4.0~mm$ voxels) \cite{Hegenscheid2013}.
The spine protocol consisted of sagittal T1- and T2-weighted turbo-spin-echo sequences with rather large slice distance only sufficient for characterizing the lumbar spine \cite{Hegenscheid2013}.

@@

%-------------------------------- Figure --------------------------------
\begin{figure*}[htb]
 \centering
 \includegraphics[width=1\textwidth, resolution=300]{figures/evaluation}
 \caption{Various case study results.
 %
 (a) Mean curvature of lumbar spine canal plotted against the mean shape of 58-74 years old female subjects (light-blue bars).
 %
 Note the high amount of this subject group in the third group. The last group only contains four outlier-subjects.
 %
 (b) Clustering of all subjects yields seven groups, where Cluster~4 assembles the mean. The light blue bars indicate the share of females in the group.
 %
 (c) A mosaic plot mapping age against dichotomous questionnaire answer to ``\emph{Did you experience back pain in the past three months?}''.
 (d) Clustering result of ``\emph{Did you experience back pain in the past three months?}'' Yes/no with female share in each group. 
 %
 Cluster 1 and 6 for answer ``\emph{Yes}'' contain mostly women.
 %
 The pivot table shows how many subjects with strong back pain are in each cluster for answer ``\emph{Yes}''.
 %
 Subjects in Cluster 1, 2 and 6 report strong back pain more often than subjects in other clusters.
%Female subjects are selected as reference for all plots, except for (c) The light blue selection in the bar charts represent the female share. }
 }
 \label{fig:application}
\end{figure*}
%-------------------------------- /Figure -------------------------------
\subsubsection{Shape Visualization and Clustering}
%
The tetrahedron-based detection model described in Section~\ref{application:Data Preprocessing} consists of corresponding grid points for each structure instance.
%
This allows for calculation of shape distance and similarity.
%
This information is used to calculate mean shapes as described in Section~\ref{Interaction- and Visualization Techniques}.

Shape distance is mapped onto color.
%
For dichotomous variables, the color codes distances between mean shapes of the two groups, for variables with more than two manifestations it encodes the distance to the global mean shape of all subjects (see Figure~\ref{fig:hypothesisfree}).
\\\\
Shape-based clustering is carried out via agglomerative hierarchical clustering of the spine canal centerlines (recall Section~\ref{application:Data Preprocessing} and \cite{Klemm2013VMV}).
%
Since the ``correct'' number of clusters in a given group is unknown, an estimate is computed by means of the knee/elbow method \cite{Salvador2004}.
%
The method has proven to produce sound results on a preliminary data set and was able to reproduce textbook knowledge \cite{Klemm2013VMV}.
\subsection{Participants, Setup and Procedure}
%
%Lam et al. divide evaluations of Visual Analysis systems in two categories, 
%
%\com{Anmerkung von Steffen einbauen: Evaluierung geht um fachliches Feedback - Wissen über epidemiologie UND rücken und wissen um die Daten - deswegen schwer ganz viele Experten ranzuziehen}
Guided by the work of Lam et al \cite{Lam2012}, we conducted an investigation of \emph{Visual Data Analysis and Reasoning} (\emph{VDAR}).
%
This approach aims to characterizes the ability of a system to explore data, discover knowledge, generate hypotheses and help formulating decisions.
%
Since it is hard to quantify when these outcomes, Lam et al. suggest case studies for the \emph{VDAR} analysis by applying the think aloud protocol to understand the domain experts observations, inferences and conclusions when using the system.

We conducted a qualitative case study with two epidemiological domain experts who also co-authored this publication.
%
Both of them are practicing physicians with focus on epidemiological research.
%
One is an specialist in internal medicine and also the head of the SHIP, the second an radiologist and responsible for the MRI-data acquisition of the study. \com{Years of experience} %and has a deep understanding on the available image data.
%
%(Their expertise and experience with the data in their day to day research make them aware of the problems associated with the analysis of medical cohort study data.)
%
\paragraph{Setup.} To bypass the large geographical distance between us and the domain experts and to take their busy schedule as physicians into account, the case study was done completely web-based.
%
%Our focus on web technologies right from the beginning made it easy to 
%Since one major reason for the focus on web technologies, no software needed to be exchanged.
The experts accessed the prototype by simply entering the website link into their browser of choice, no exchange of software was required.
%Since our prototype ran on a local server, simply needed the homepage link to access the prototype.
%
%A website link was send to the users, which they could open with their browser of choice and start using the tool.
%
To make the user input comprehensible, we observed their monitor via screen sharing.
%
Communication was enabled via webcam supported voice over ip.
%
The total setup time including installing the screen sharing application was about five minutes.
%
Video-recordings of the sessions allowed a detailed evaluation afterwards.%made a possible to evaluate the sessions later on.
%
\paragraph{Procedure.}
At first, we controlled mouse and keyboard of the participants pc and demonstrated the basic functionalities of the system.
%
%This included the contingency matrix, the correlation view, how to introduce feature visualizations, how the color coding of differences works and the pivot tables.
%
As they understood the concepts, we handed over the mouse and keyboard control and only observed from this point on.
%
The epidemiologists were given two tasks: one hypothesis-free analysis of the data and one starting with an assumption.
%
For each case we conducted one analysis with each expert.

% \subsection{Exploratory Analysis of the Lumbar Spine Data Set}
\subsection{Case 1: Hypothesis-free Analysis} \label{Hypothesis-free analysis}
Analyzing the data set without prior hypothesis requires a starting point giving an overview over the data first \cite{Shneiderman1996}.
%
With our tool, there are two ways to achieve this.
%
Performing a \emph{local investigation} by clustering all subjects or use the contingency matrix for a \emph{multivariate analysis}.
%
The latter was chosen first by both domain experts.
% The latter was chosen first by the domain expert with radiological background.
%
%They pointed out the high potential they see in this visualization, because it allows to look at the data in an unbiased way.
%They pointed out the benefits of the possibility to look at all variables in the context of each other.
Before, they were not able to to look at all variables in the context of each other.
%
%
To cite one expert, the contingency matrix ``illumunates the data black box'', making it possible to look at the data unbiased from assumptions.

\paragraph{Analysis 1.}
%
The radiologist was looking for correlations with shape related features in the data, finding that they correlate with \emph{leg pain}, \emph{age}, \emph{body height} and \emph{hormone replacement therapy status}.
%
Due to the dense mapping of information in the contingency matrix it was suggested to make this visualization full screen.
%Many correlations observed in this view

After this initial overview, the expert performed a \emph{multivariate analysis} by introducing features, such as \emph{age}, \emph{waist circumference}, \emph{weight} or \emph{lumbar spine canal curvature} as bar charts views into the canvas area and selected subgroups to see how they are distributed and if they could observe unusual behavior in the mean-shapes.
%
This pointed out problems with the used categorization method splitting numerical variables into equally sized ordinal bins.
%
If an feature contains outliers, such as \emph{waist circumference} (e.g. by subjects with morbid obesity) this approach leads to sparse categories, making it hard to calculate associations.
%
The proposed expert solution for this is categorization using quantiles/quartiles and is described in detail in Section~\ref{Lessons Learned}.

A \emph{multivariate analysis} using the \emph{Cram\'{e}r's V} contingency values for subjects with strong lumbar spine curvature showed, that these subjects are prominetly females between 58-74 years who also report pain radiating from their back into other body regions Fig.~\ref{fig:application}(a).
%
%\com{Allgemein: Workflow noch einmal grob beschreiben - Overview first, details on demand}
%
\paragraph{Analysis 2.}
% \begin{figure}[htb]
%  \centering
%  %\includegraphics[width=1.5in]{figures/InteractionLoop}
%  \includegraphics[width=3.5in]{figures/hypothesisfree}
%  \caption{(a) Clustering result of all subjects. The bar chart height indicates the number of subjects in the cluster. The difference to the mean shape is color-coded, whereas blue represents no difference and red a large difference. (b) A pivot table relates cluster results to gender and employment status. Cluster 8 is striking because of the high proportion of women, while cluster 2, 3, 4 contain almost only men.}
%  \label{fig:hypothesisfree}
% \end{figure}
%
The second expert started also with a \emph{multivariate analysis} using the contingency matrix to analyze non-image-derived features, such as age-associated parameters like \emph{income}, \emph{blood fat values} or \emph{number of born children}, but found no associations of interest.
%
Therefore, he applied the \emph{local investigation} pattern by a shape grouping step using shape-based clustering via dragging the \emph{All subjects} from the sidebar into the canvas area, triggering the shape clustering.
%Shape-based hypothesis-free exploration starts with the \emph{local investigation} pattern by a shape grouping step using shape-based clustering.
%
The results for this step can be seen in Figure~\ref{fig:application} (b).

Cluster 4 represents subjects with average shape.
%
Other shapes differ with respect to size, such as cluster 2, 3, 7, where the last one and cluster 5 also represent a more straight spine, which is usual for subjects with larger body size.
%
Cluster 1 and 6 contain outliers, characterized by their unusual shape and small number.
%
%Cluster 8 has the second largest number of elements and was therefore of special interest.
%
%To perform an automatic feature suggestion using a \emph{multivariate analysis} by looking at \emph{Cram\'{e}r's V} contingency values of in the contingency pane.
To get an overview over the suggested features, the user opened the contingency pane to perform a \emph{multivariate analysis} by looking at \emph{Cram\'{e}r's V} contingency values of all clusters, revealing a strong correlation with \emph{gender} and \emph{body size}.
%
Therefore another \emph{multivariate analysis} was carried out by the feature \emph{gender} to the canvas from the contingency pane and selecting all female subjects (Figure~\ref{fig:application} (b)).
%
Cluster 1 contained almost only female subjects.
%
Contingency values for this cluster revealed correlations with \emph{leg fatigue}, \emph{physical heavy work}, \emph{body weight}, \emph{dyspnoea} and \emph{headache intensity}.
%
%This revealed associations of this group with employment status, gender, body size, age, thyroid nodules and blood fat value.
%
Since it is an pain indicator, headache was of special interest and was further investigated by incorporating a pivot table setting \emph{headache intensity} in relation to cluster affiliation.
%
It was found, that cluster~1 subjects report heavy headaches more frequently than other subjects.
%Another \emph{multivariate analysis} using a pivot table set gender and employment status in relation to clustering affiliation\add{.} 
%
%\add{It s}hows that cluster~8 contains mostly women and also has a larger unemployment rate (see Figure~\ref{fig:hypothesisfree}), while the overall employment rate of women and men in the data set is almost exactly 50\%.
%
%While all observed features seem to be plausible associations related to back pain, the values indicate that cluster~8 contains subjects with chronic back pain radiating to the legs.
%
%Metabolic parameters, such as blood fat and blood sugar, are also possibly associated features.
%
%The employment status is a feature relating to many different lifestyle factors such as income or nutrition as well as age and might act as confounder.
\\\\
The experts emphasized the importance of methods providing overview over the data for hypothesis generation.
%
With the presented \emph{IVA} approaches they were quickly able to confirm medical knowledge as well as elaborate new hypothesis.
%
We observed, that even though the visualization may treat every feature equally, the domain experts is more likely interested in features they are familiar with and have personal clinical experience with.
% TODO: Contingency Matrix or Adjecency Matrix
%\com{00.44 Clustering-Ergebnis Weiblich und Arbeitspensum Alter}

\subsection{Case 2: Hypothesis-based Analysis}
If the user proposes a hypothesis about a relation between a non-image feature regarding shape, the workflow slightly differs from the hypothesis-free analysis.
%
The starting point follows the \emph{feature investigation} pattern, where a feature of interest is selected by dragging it into the canvas area and viewing the subject's distribution as well as their shape differences.
% \paragraph{Analysis 1.}
% The radiologist performed a \emph{feature investigation} by dragging the age variable into the canvas area, observing that age-bins 36-52 years and 52-68 years are nearly identical to the global mean, while 20-36 years and 68-84 years differ in size.
% %
% Performing a \emph{multivariate analysis} by selecting only female subjects via the additionally added gender feature it was observed, that distance was closest to subjects of 68-84 years old and largest to 20-36 years old.
% %
% The expert argued that this is most likely due to the shrinking of the spine with increasing age, making it more similar to the female average spine which is smaller than the global average \com{(00.29)}.
% %
% 
\paragraph{Analysis 1.}
Hypothesis: ``\emph{Back pain is associated with age and lumbar spine shape}''.
%Hypothesis: "Back pain is associated with age and therefore there differences in the lumbar spine shape are observable." 
%
To validate this hypothesis, a \emph{feature investigation} was performed by introducing the dichotomous feature ``\emph{Did you experience back pain in the last three month?}'' together with the age as mosaic plot by dropping both features on the canvas area.
%
The expert was not able to observe the expected effect in the visualization.
%
Reasons for this are twofold.
%
For one, age influences the lumbar spine shape, as seen in Fig~\ref{fig:application}~(c), while the differences between subjects with and without back pain are small.
%
The major difference seen in the visualization are therefore presented by the age feature, masking differences via the back pain parameter.
%
The second explanation is the commonality of back pain in our society.
%
As seen in Fig~\ref{fig:application}~(c), subjects reporting back pain are even the majority, which makes it difficult to extract parameters which reliably describe back pain.
%
A \emph{multivariate analysis} using the contingency table showed a strong association between \emph{back pain} with \emph{gender} and \emph{body height}.
%
\emph{Body height} was explained to be a confounder for \emph{gender}, since female subjects smaller on average than male subjects.
%
The analysis solely based on shape accentuated differences in \emph{gender}, not \emph{back pain}.
%This makes it hard to analyze shape related features for this disease
%
The next step is to use a variable describing the strength of back pain and only analyzing strong back pain!

The epidemiologists pointed out that they would like to see a more intuitive and fast way to select subgroups based on different features to make full use of the analysis capabilities, as discussed in Section~\ref{Lessons Learned}.
%gender, use of pain medications. this would be good for future work, providing a better way to subdivide the data set and construct groups. The presented method is very good to gain an overview

@@

% ---------------------------------------------------------------------------------------------------------
\subsection{Further Feedback and Lessons Learned} \label{Lessons Learned}
%\com{ToDo: Lessons Learned der einzelnen Cases - Fokus bei Hypothesenfreier Analyse muss etwa eine gute Übersichtsvisualisierung sein, da es genau das ist was den Epidemiologen hierzu im Moment fehlt}
%
Both domain experts concluded positively over the benefits of the \emph{IVA} approach when analyzing image based cohort study data.
%
One expert emphasized the way the image data is included into information visualizations which comes much more natural to her due to her background in radiology.
%
%Both experts quickly understood the basic concepts of the visualizations and were able to 
%
Great potential is also seen in communicating insights efficiently using the presented visualizations \com{(1.08)}.
%
\paragraph{Multivariate analysis is most important for hypothesis generation.}
Both experts emphasized the potential of the \emph{multivariate analysis} capabilities of the adjacency matrix for getting insight into a large amount of features simultaneously.
%
It is also useful to verify established but still controversial risk factors, such as the metabolic syndrome for coronary heart disease and whether the data set provides even more suitable risk factors \com{(39.32)}.
%
Creating adjacency matrices for sub groups, such as different age bins can help to characterize the aging process by deriving age-specific risk factors.

@@

The experts also emphasized that associations related to shape strongly depend on the segmentation quality.
%
The lumbar spine model used in this case study captures deformation of the spine canal well, but lacks precise definition in vertebrae height and shape.
%
Since deformation of the spine canal is the last stage of pathological lumbar spine deformation and is preceded by vertebrae deformation, the system would strongly benefit from more precise segmentation results capturing these prior changes.
%
For the visual comparison itself, the radiologist proposed an abstraction the representation them into representative land marks, such as centers of the vertebrae and cardinal points of the lumbar spine canal.
